---
title: splines2
output: github_document
---

[![CRAN_Status_Badge][r-pkg-badge]][cran-url]
[![Downloads from the RStudio CRAN mirror][cranlog-badge]][cran-url]
[![Build Status][gha-icon]][gha-url]
[![codecov][codecov-main]][codecov]


```{r setup, echo = FALSE}
knitr::opts_chunk$set(comment = "")
options(width = 100, digits = 5)
```

The R package **splines2** provides functions to construct basis matrix of

- B-splines
- M-splines
- I-splines
- convex splines (C-splines)
- generalized Bernstein polynomials
- their integrals (except C-splines) and derivatives of given order by
  close-form recursive formulas

In addition to the R interface, **splines2** also provides a C++ header-only
library integrated with **Rcpp**, which allows construction of spline basis
matrix directly in C++ with the help of **Rcpp** and **RcppArmadillo**.  So it
can also be treated as one of the **Rcpp\*** packages.  A toy example package
that uses the C++ interface is available [here][example-pkg].


## Installation of CRAN Version

You can install the released version from [CRAN][cran-url].

```R
install.packages("splines2")
```

## Development

The latest version of package is under development at [GitHub][github-url].  If
it is able to pass the automated package checks, one may install it by

```R
if (! require(remotes)) install.packages("remotes")
remotes::install_github("wenjie2wang/splines2", upgrade = "never")
```

## Getting Started

[Online document][pkgdown-url] provides reference for all functions and contains
the following vignettes:

- [Demonstration of the common usages in R through examples][pkg-demo].
- [Introduction to the usage with Rcpp][pkg-wi-rcpp]

## Performance

Since v0.3.0, the implementation of the main functions has been rewritten in C++
with the help of the **Rcpp** and **RcppArmadillo** package. The computational
performance has thus been boosted.

Some benchmarks with the **splines** package
(version `r packageVersion("splines")`)
are provided for reference as follows:

```{r bench-setup}
library(microbenchmark)
library(splines)
library(splines2)

x <- seq.int(0, 1, 0.001)
degree <- 3
ord <- degree + 1
knots <- seq.int(0.1, 0.9, 0.1)
b_knots <- range(x)
all_knots <- sort(c(knots, rep(b_knots, ord)))

## check equivalency of outputs
my_check <- function(values) {
    all(sapply(values[- 1], function(x) {
        all.equal(unclass(values[[1]]), x, check.attributes = FALSE)
    }))
}
```

For B-splines, function `splines2::bSpline()` provides equivalent results with
`splines::bs()` and `splines::splineDesign()`, and is about 3x faster than
`bs()` and 2x faster than `splineDesign()`.

```{r bench-bSpline}
## B-splines
microbenchmark(
    "splines::bs" = bs(x, knots = knots, degree = degree,
                       intercept = TRUE, Boundary.knots = b_knots),
    "splines::splineDesign" = splineDesign(x, knots = all_knots, ord = ord),
    "splines2::bSpline" = bSpline(x, knots = knots, degree = degree,
                                  intercept = TRUE, Boundary.knots = b_knots),
    check = my_check,
    times = 1e3
)
```

Similarly, for derivatives of B-splines, `splines2::dbs()` provides equivalent
results with `splines::splineDesign()`, and is more than 2x faster.

```{r bench-dbs}
## Derivatives of B-splines
derivs <- 2
microbenchmark(
    "splines::splineDesign" = splineDesign(x, knots = all_knots,
                                           ord = ord, derivs = derivs),
    "splines2::dbs" = dbs(x, derivs = derivs, knots = knots, degree = degree,
                          intercept = TRUE, Boundary.knots = b_knots),
    check = my_check,
    times = 1e3
)
```

The **splines** package does not provide function producing integrals of
B-splines.  So we instead performed a comparison with package **ibs** (version
`r packageVersion("ibs")`), where the function `ibs::ibs()` was also implemented
in **Rcpp**.

```{r bench-ibs}
## integrals of B-splines
set.seed(123)
coef_sp <- rnorm(length(all_knots) - ord)
microbenchmark(
    "ibs::ibs" = ibs::ibs(x, knots = all_knots, ord = ord, coef = coef_sp),
    "splines2::ibs" = as.numeric(
        splines2::ibs(x, knots = knots, degree = degree,
                      intercept = TRUE, Boundary.knots = b_knots) %*% coef_sp
    ),
    check = my_check,
    times = 1e3
)
```

The function `ibs::ibs()` returns the integrated B-splines instead of the
integrals of spline bases.  So we applied the same coefficients to the bases
from `splines2::ibs()` for equivalent results, which was still much faster than
`ibs::ibs()`.


## License

[GNU General Public License][gpl] (≥ 3)

[r-pkg-badge]: https://www.r-pkg.org/badges/version/splines2
[cranlog-badge]: https://cranlogs.r-pkg.org/badges/splines2
[cran-url]: https://CRAN.R-project.org/package=splines2
[gha-icon]: https://github.com/wenjie2wang/splines2/workflows/R-CMD-check/badge.svg
[gha-url]: https://github.com/wenjie2wang/splines2/actions
[github-url]: https://github.com/wenjie2wang/splines2
[pkgdown-url]: https://wwenjie.org/splines2
[example-pkg]: https://github.com/wenjie2wang/example-pkg-Rcpp-splines2
[pkg-demo]: https://wwenjie.org/splines2/articles/splines2-intro
[pkg-wi-rcpp]: https://wwenjie.org/splines2/articles/splines2-wi-rcpp
[codecov]: https://codecov.io/gh/wenjie2wang/splines2
[codecov-main]: https://codecov.io/gh/wenjie2wang/splines2/branch/main/graph/badge.svg
[gpl]: https://www.gnu.org/licenses/

