---
title: splines2
output: github_document
---

[![CRAN_Status_Badge][r-pkg-badge]][cran-url]
[![Downloads from the RStudio CRAN mirror][cranlog-badge]][cran-url]
[![Build Status][travis-master]][travis]
[![AppVeyor Build Status][appveyor-master]][appveyor]
[![codecov][codecov-master]][codecov]


```{r setup, echo = FALSE}
knitr::opts_chunk$set(comment = "")
options(width = 100, digits = 5)
```

The R package **splines2** is a supplementary package on splines providing
functions constructing B-splines, integral of B-splines, monotone splines
(M-splines) and its integral (I-splines), convex splines (C-splines), and their
derivatives of given order.


## Installation of CRAN Version

You can install the released version from [CRAN][cran-url].

```R
install.packages("splines2")
```

## Development

The latest version of package is under development at [GitHub][github-url].  If
it is able to pass the building check by Travis CI, one may install it by

```R
if (! require(remotes)) install.packages("remotes")
remotes::install_github("wenjie2wang/splines2")
```

## Getting Started

- [Online documentation][pkgdown-url] provides reference for all functions
- Package vignette for a quick [demonstration][pkg-demo] for the common usage
  through examples

## Performance

Since v0.3.0, the implementation of the main functions has been rewritten in C++
with the help of the **Rcpp** and **RcppArmadillo** package. The computational
performance has thus been boosted.

Some benchmarks with the **splines** package
(version `r packageVersion("splines")`)
are provided for reference as follows:

```{r bench-setup}
library(microbenchmark)
library(splines)
library(splines2)

x <- seq.int(0, 1, 0.001)
degree <- 3
ord <- degree + 1
knots <- seq.int(0.1, 0.9, 0.1)
b_knots <- c(0, 1)
all_knots <- sort(c(knots, rep(range(x), degree + 1)))

## check equivalency of outputs
my_check <- function(values) {
    all(sapply(values[- 1], function(x) {
        all.equal(unclass(values[[1]]), x, check.attributes = FALSE)
    }))
}
```

For B-splines, function `splines2::bSpline()` provides equivalent results with
`splines::bs()` and `splines::splineDesign()`, and is about 3x faster than
`bs()` and 2x faster than `splineDesign()`.

```{r bench-bSpline}
## B-splines
microbenchmark(
    "splines::bs" = bs(x, knots = knots, degree = degree,
                       intercept = TRUE, Boundary.knots = b_knots),
    "splines::splineDesign" = splineDesign(x, knots = all_knots, ord = ord),
    "splines2::bSpline" = bSpline(x, knots = knots, degree = degree,
                                  intercept = TRUE, Boundary.knots = b_knots),
    check = my_check,
    times = 1e3
)
```

Similarly, for derivatives of B-splines, `splines2::dbs()` provides equivalent
results with `splines::splineDesign()`, and is more than 2x faster.

```{r bench-dbs}
## Derivatives of B-splines
derivs <- 2
microbenchmark(
    "splines::splineDesign" = splineDesign(x, knots = all_knots,
                                           ord = ord, derivs = derivs),
    "splines2::dbs" = dbs(x, derivs = derivs, knots = knots, degree = degree,
                          intercept = TRUE, Boundary.knots = b_knots),
    check = my_check,
    times = 1e3
)
```

The **splines** package does not provide function producing integrals of
B-splines.  So we instead performed a comparison with package **ibs** (version
`r packageVersion("ibs")`), where the function `ibs::ibs()` was also implemented
in **Rcpp**.

```{r bench-ibs}
## integrals of B-splines
set.seed(123)
coef_sp <- rnorm(length(all_knots) - ord)
microbenchmark(
    "ibs::ibs" = ibs::ibs(x, knots = all_knots, ord = ord, coef = coef_sp),
    "splines2::ibs" = as.numeric(
        splines2::ibs(x, knots = knots, degree = degree,
                      intercept = TRUE, Boundary.knots = b_knots) %*% coef_sp
    ),
    check = my_check,
    times = 1e3
)
```

The function `ibs::ibs()` returns the integrated B-splines instead of the
integrals of spline bases.  So we applied the same coefficients to the bases
from `splines2::ibs()` for equivalent results, which was still much faster than
`ibs::ibs()`.


## License

[GNU General Public License (v3)][gpl]

[r-pkg-badge]: https://www.r-pkg.org/badges/version/splines2
[cranlog-badge]: https://cranlogs.r-pkg.org/badges/splines2
[cran-url]: https://CRAN.R-project.org/package=splines2
[travis]: https://travis-ci.org/wenjie2wang/splines2
[travis-master]: https://travis-ci.org/wenjie2wang/splines2.svg?branch=master
[github-url]: https://github.com/wenjie2wang/splines2
[pkgdown-url]: https://wenjie-stat.me/splines2
[pkg-demo]: https://wenjie-stat.me/splines2/articles/splines2-intro
[codecov]: https://codecov.io/gh/wenjie2wang/splines2
[codecov-master]: https://codecov.io/gh/wenjie2wang/splines2/branch/master/graph/badge.svg
[appveyor]: https://ci.appveyor.com/project/wenjie2wang/splines2
[appveyor-master]: https://ci.appveyor.com/api/projects/status/bvoso7nxchg1incb/branch/master?svg=true
[gpl]: https://www.gnu.org/licenses/
