---
title: "Introduction to splines2"
author: Wenjie Wang
date: "`r Sys.Date()`"
bibliography:
- ../inst/bib/splines2.bib
vignette: >
  %\VignetteIndexEntry{Introduction to splines2}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
output: rmarkdown::html_vignette
---


The package **splines2** is now designed to be a complementary package on
splines.  It provides functions constructing a variety of spline bases that are
not available from the package **splines** shipped with **R**.  Most functions
in **splines2** have a very similar user interface with the function `bs` in
package **splines**. Currently, **splines2** provides function constructing
B-splines, monotone splines (M-splines) and its integral (I-splines), convex
splines (C-splines), and integral of B-splines.  Compared with package
**splines**, piecewise constant basis is allowed for B-spline and M-spline basis
in **splines2**.


In this vignette, we introduce the basic usage of the functions provided by
examples. The details of function syntax and attributes bundled with objects
returned are available in the package manual.


An outline of the remainder of the vignette is as follows: We first introduce
the functions constructing the monotone splines ([M-splines](#mSpline)), its
integral ([I-splines](#iSpline)), and convex splines ([C-splines](#cSpline)).
After then, the examples for [integral of B-splines](#ibs) are provided, which
include one toy example for integral of degree zero or piecewise constant
B-spline.  Last but not the least, handy methods of S3 generic
function [predict](#predict) for objects produced by **splines2** are
demonstrated for the evaluation of the same spline basis at new values.


## M-splines using `mSpline`{#mSpline}

M-splines [@ramsay1988monotone] can be considered as a normalized version of
B-splines with unit integral within boundary knots.  An example given by
@ramsay1988monotone was a quadratic M-splines with three internal knots placed
at 0.3, 0.5, and 0.6. The boundary knots by default are the range of the data
`x`. So in this example, they are 0 and 1.


```{r mSpline, fig.width=7, fig.height=4}
library(splines2)
knots <- c(0.3, 0.5, 0.6)
x <- seq(0, 1, 0.01)
msOut <- mSpline(x, knots = knots, degree = 2, intercept = TRUE) # M-splines
library(graphics) # attach graphics just in case for plots
par(mar = c(2.5, 2.5, 0, 0), mgp = c(1.5, 0.5, 0))
matplot(x, msOut, type = "l", ylab = "Quadratic M-splines")
abline(v = knots, lty = 2, col = "gray") # internal knots
```


## I-splines using `iSpline` {#iSpline}

I-splines [@ramsay1988monotone] are simply the integral of M-splines and thus
monotonically non-decreasing with unit maximum value. A monotonically
non-decreasing (non-increasing) function can be fitted by a linear combination
of I-spline bases with non-negative (non-positive) coefficients, plus a constant
function (where the coefficient of the constant function is unconstrained).

The example given by @ramsay1988monotone was the I-splines corresponding to that
quadratic M-splines with three internal knots placed at 0.3, 0.5, and 0.6. Note
that the degree of I-splines is defined by the associated M-splines instead of
their polynomial degree.


```{r iSpline, fig.width=7, fig.height=4}
isOut <- iSpline(x, knots = knots, degree = 2, intercept = TRUE) # I-splines
par(mar = c(2.5, 2.5, 0, 0), mgp = c(1.5, 0.5, 0))
matplot(x, isOut, type = "l", ylab = "I-splines of degree two")
abline(h = 1, v = knots, lty = 2, col = "gray") # internal knots
```

In additional, the corresponding M-spline basis matrix is shipped in the
attribute named `msMat` in the object returned by function `iSpline`. In other
words, if we need both M-spline bases and their integral splines in model
fitting, extra function call on `mSpline` can be avoided conveniently.

```{r msMat}
all.equal(msOut, attr(isOut, "msMat")) # 'msMat' attribute for M-splines
```


## C-splines using `cSpline` {#cSpline}

Convex splines [@meyer2008inference] called C-splines can be considered as a
scaled version of the integral of I-splines with unit maximum value.
@meyer2008inference applied C-splines to shape-restricted regression
analysis. The monotone property of I-spines ensures the convexity of
C-splines. A convex regression function can be estimated using linear
combinations of the C-spline bases with non-negative coefficients, plus an
unrestricted linear combination of the constant function and the identity
function $g(x) = x$. If the underlying regression function is both increasing
and convex, the coefficient on the identity function is restricted to be
nonnegative as well.


```{r cSpline, fig.width=7, fig.height=4}
cMat <- cSpline(x, knots = knots, degree = 2, intercept = TRUE)
par(mar = c(2.5, 2.5, 0, 0), mgp = c(1.5, 0.5, 0))
matplot(x, cMat, type = "l", ylab = "C-spline basis")
abline(h = 1, v = knots, lty = 2, col = "gray")
```

<!-- FIXME -->

## Integral B-splines using `ibs` {#ibs}

Formula from @de1978practical.

```{r intBs1, fig.width=7, fig.height=4}
knots <- c(0.2, 0.4, 0.7, 0.9)
ibsOut <- ibs(x, knots = knots, degree = 1, intercept = TRUE)
par(mar = c(2.5, 2.5, 0, 0), mgp = c(1.5, 0.5, 0))
matplot(x, ibsOut, type = "l", ylab = "Integral of B-spline bases")
abline(v = knots, lty = 2, col = "gray")
matplot(x, attr(ibsOut, "bsMat"), type = "l",
        ylab = "B-spline bases")
abline(v = knots, lty = 2, col = "gray")
```


## Reference

