---
title: "Introduction to splines2"
author: Wenjie Wang
date: "`r Sys.Date()`"
bibliography:
- ../inst/bib/splines2.bib
vignette: >
  %\VignetteIndexEntry{Introduction to splines2}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
output: rmarkdown::html_vignette
---

```{r setup, echo = FALSE}
knitr::opts_knit$set(global.par = TRUE)
knitr::opts_chunk$set(fig.width = 7, fig.height = 4)
```

```{r set-par, echo = FALSE}
library(graphics)
par(mar = c(2.5, 2.5, 0.2, 0.1), mgp = c(1.5, 0.5, 0))
```

The **splines2** package provides functions constructing a variety of regression
spline bases that are not available from the **splines** package shipped with
base R.  Most functions have a very similar user interface with the function
`splines::bs()`.  To be more specific, it provides function constructing
B-splines, integral of B-splines, M-splines and its integral (I-splines), convex
splines (C-splines), generalized Bernstein polynomials, and their derivatives of
given order.  Compared with the **splines** package, **splines2** allows
piece-wise constant basis for B-splines and provides a more user-friendly
interface for their derivatives with consistent handling on `NA`'s.  Most of the
implementations has been (re)written in C++ with the help of **Rcpp** and
**RcppArmadillo** since v0.3.0, which helps boost the computational performance.


In the remaining of this vignette, we introduce the basic usage of most
functions in the package through examples. The details of function syntax are
available in the package manual and thus will be not discussed.


## B-splines with their integrals and derivatives {#bSpline}

Function `bSpline()` provides B-spline basis matrix and allows `degree = 0` for
piece-wise constant bases, which extends the `bs()` function in package
**splines** with a better computational performance.  One example of linear
B-splines with two internal knots is given as follows:

```{r bSpline, fig.cap="B-splines of degree one with two internal knots."}
library(splines2)
knots <- c(0.3, 0.5, 0.6)
x <- seq(0, 1, 0.01)
bsMat <- bSpline(x, knots = knots, degree = 1, intercept = TRUE)
matplot(x, bsMat, type = "l", ylab = "y")
abline(v = knots, lty = 2, col = "gray")
```

The close-form recursive formula of B-spline integrals and derivatives given by
@de1978practical is implemented in function `ibs()` and `dbs()`, respectively.
Two toy examples are given as follows:

```{r ibs, fig.cap="Piecewise linear B-splines (left) and their integrals (right)."}
ibsMat <- ibs(x, knots = knots, degree = 1, intercept = TRUE)
par(mfrow = c(1, 2))
matplot(x, bsMat, type = "l", ylab = "y")
abline(v = knots, h = 1, lty = 2, col = "gray")
matplot(x, ibsMat, type = "l", ylab = "y")
abline(v = knots, h = c(0.15, 0.2, 0.25), lty = 2, col = "gray")
```

```{r dbs, fig.cap="Cubic B-splines (left) and their first derivative (right)."}
bsMat <- bSpline(x, knots = knots, intercept = TRUE)
dbsMat <- dbs(x, knots = knots, intercept = TRUE)
par(mfrow = c(1, 2))
matplot(x, bsMat, type = "l", ylab = "y")
abline(v = knots, lty = 2, col = "gray")
matplot(x, dbsMat, type = "l", ylab = "y")
abline(v = knots, lty = 2, col = "gray")
```

We may also obtain the derivatives easily by the `deriv()` method as follows:

```{r dbsMat}
is_equivalent <- function(a, b) {
    all.equal(a, b, check.attributes = FALSE)
}
stopifnot(is_equivalent(dbsMat, deriv(bsMat)))
```

## M-splines using `mSpline()` {#mSpline}

M-splines [@ramsay1988monotone] can be considered as a normalized version of
B-splines with unit integral within boundary knots.  An example given by
@ramsay1988monotone was a quadratic M-splines with three internal knots placed
at 0.3, 0.5, and 0.6. The boundary knots by default are the range of the data
`x`, thus 0 and 1 in this example.

```{r reset-par-mSpline, echo = FALSE}
par(mfrow = c(1, 1))
```

```{r mSpline, fig.cap = "Quadratic M-splines with three internal knots."}
msMat <- mSpline(x, knots = knots, degree = 2, intercept = TRUE)
matplot(x, msMat, type = "l", ylab = "y")
abline(v = knots, lty = 2, col = "gray")
```

The derivative of given order of M-splines can be obtained by specifying a
positive integer to argument `dervis` of `mSpline()`.  Also, for an existing
`mSpline` object generated by `mSpline()`, the `deriv()` method can be used
conveniently.  For example, the first derivative of the M-splines given in last
example can be obtained equivalently as follows:

```{r mSpline-derivs}
dmsMat1 <- mSpline(x, knots = knots, degree = 2, intercept = TRUE, derivs = 1)
dmsMat2 <- deriv(msMat)
stopifnot(is_equivalent(dmsMat1, dmsMat2))
```

## I-splines using `iSpline()` {#iSpline}

I-splines [@ramsay1988monotone] are simply the integral of M-splines and thus
monotonically non-decreasing with unit maximum value. A monotonically
non-decreasing (non-increasing) function can be fitted by a linear combination
of I-spline bases with non-negative (non-positive) coefficients *plus a
constant*, where the coefficient of the constant is unconstrained.

The example given by @ramsay1988monotone was the I-splines corresponding to that
quadratic M-splines with three internal knots placed at 0.3, 0.5, and
0.6. Notice that the degree of I-splines is defined from the associated
M-splines instead of their own polynomial degree.

```{r iSpline, fig.cap = "I-splines of degree two with three internal knots."}
isMat <- iSpline(x, knots = knots, degree = 2, intercept = TRUE)
matplot(x, isMat, type = "l", ylab = "y")
abline(h = 1, v = knots, lty = 2, col = "gray")
```

The corresponding M-spline basis matrix can be obtained easily as the first
derivatives of the I-splines by the `deriv()` method.

```{r msMat}
stopifnot(is_equivalent(msMat, deriv(isMat)))
```

We may specify the `derivs = 2` in the `deriv()` method for the second
derivatives of the I-splines, which are equivalent to the first derivatives of
the corresponding M-splines.

```{r dmsMat}
dmsMat3 <- deriv(isMat, 2)
stopifnot(is_equivalent(dmsMat1, dmsMat3))
```

## C-splines using `cSpline` {#cSpline}

Convex splines [@meyer2008inference] called C-splines are scaled integrals of
I-splines with unit maximum value at the right boundary knot.
@meyer2008inference applied C-splines to shape-restricted regression
analysis. The monotone (non-decreasing) property of I-spines ensures the
convexity of C-splines. A convex regression function can be estimated using
linear combinations of the C-spline bases with non-negative coefficients, plus
an unconstrained linear combination of a constant and an identity function
$g(x)=x$. If the underlying regression function is both increasing and convex,
the coefficient on the identity function is restricted to be non-negative as
well.


We may specify the argument `scale = FALSE` in function `cSpline()` to disable
the scaling of the integrals of I-splines.  Then the actual integrals of the
corresponding I-splines will be returned.  If `scale = TRUE` (by default), each
C-spline basis is scaled to have unit height at the right boundary knot.


```{r cSpline-scaled, fig.cap = "C-splines of degree two with three internal knots."}
csMat1 <- cSpline(x, knots = knots, degree = 2, intercept = TRUE)
matplot(x, csMat1, type = "l", ylab = "y")
abline(h = 1, v = knots, lty = 2, col = "gray")
```

Similarly, the `deriv()` method can be used to obtain the derivatives.  A nested
call of `deriv()` is supported for derivatives of order greater than one.
However, the argument `derivs` of the `deriv()` method can be specified directly
for a better computational performance.  For example, the first and second
derivatives can be obtained by the following equivalent approaches,
respectively.

```{r cSpline-not-scaled}
csMat2 <- cSpline(x, knots = knots, degree = 2, intercept = TRUE, scale = FALSE)
stopifnot(is_equivalent(isMat, deriv(csMat2)))
stopifnot(is_equivalent(msMat, deriv(csMat2, 2)))
stopifnot(is_equivalent(msMat, deriv(deriv(csMat2))))
```

## Generalized Bernstein Polynomials

The Bernstein polynomials has also been applied to shape-constrained regression
analysis [@wang2012csda].  The $i$-th basis of the generalized Bernstein
polynomials of degree $n$ over $[a, b]$ is defined as follows:
$$
B_i^n(x)=\frac{1}{(b-a)^n}{n\choose i}(x-a)^i (b-x)^{n-i},~0\le i\le n,
$$
where $a\le x\le b$.
Obviously, it reduces to Bernstein polynomials defined over $[0, 1]$ when $a =
0$ and $b = 1$.

We may obtain the basis matrix of the generalized using the function
`bernsteinPoly()`.  For example, the Bernstein polynomials of degree 4 over [0,
1] and is generated as follows:

```{r bp-1, fig.cap="Bernstein polynomials of degree 4 over [0, 1] (left) and the generalized version over [- 1, 1] (right)."}
x1 <- seq.int(0, 1, 0.01)
x2 <- seq.int(- 1, 1, 0.01)
bpMat1 <- bernsteinPoly(x1, degree = 4, intercept = TRUE,
                        Boundary.knots = c(0, 1))
bpMat2 <- bernsteinPoly(x2, degree = 4, intercept = TRUE,
                        Boundary.knots = c(- 1, 1))
par(mfrow = c(1, 2))
matplot(x1, bpMat1, type = "l", ylab = "y")
matplot(x2, bpMat2, type = "l", ylab = "y")
```

In addition, we may specify `integral = TRUE` or `derivs = 1` in
`bernsteinPoly()` for their integrals or first derivatives, respectively.

```{r bp-2, fig.height=6}
ibpMat1 <- bernsteinPoly(x1, degree = 4, intercept = TRUE, integral = TRUE)
ibpMat2 <- bernsteinPoly(x2, degree = 4, intercept = TRUE, integral = TRUE,
                         Boundary.knots = c(- 1, 1))
dbpMat1 <- bernsteinPoly(x1, degree = 4, intercept = TRUE, derivs = 1)
dbpMat2 <- bernsteinPoly(x2, degree = 4, intercept = TRUE, derivs = 1,
                         Boundary.knots = c(- 1, 1))
par(mfrow = c(2, 2))
matplot(x1, ibpMat1, type = "l", ylab = "y")
matplot(x2, ibpMat2, type = "l", ylab = "y")
matplot(x1, dbpMat1, type = "l", ylab = "y")
matplot(x2, dbpMat2, type = "l", ylab = "y")
```

Similarly, we may also use the `deriv()` method to get derivatives of an
existing `bernsteinPoly` object.

```{r bp-deriv}
stopifnot(is_equivalent(dbpMat1, deriv(bpMat1)))
stopifnot(is_equivalent(dbpMat2, deriv(bpMat2)))
stopifnot(is_equivalent(dbpMat1, deriv(ibpMat1, 2)))
stopifnot(is_equivalent(dbpMat2, deriv(ibpMat2, 2)))
```

## Evaluation on New Values by `predict` {#predict}

The methods for **splines2** objects dispatched by generic function `predict`
will be useful if we want to evaluate the spline object at possibly new $x$
values. For instance, we may evaluate the value of I-splines object in previous
example at 0.275, 0.525, and 0.8, respectively, as follows:

```{r predict}
new_x <- c(0.275, 0.525, 0.8)
names(new_x) <- paste0("x=", new_x)
predict(isMat, new_x)
```

Technically speaking, the methods take all information needed, such as `knots`,
`degree`, `intercept`, etc., from attributes of the original objects and call
the corresponding function automatically for those new $x$ values. Therefore,
the `predict` methods will not be applicable if those attributes are somehow
lost after some operations.



## Reference

