---
title: "Introduction to splines2"
author: Wenjie Wang
date: "`r Sys.Date()`"
bibliography:
- ../inst/bib/splines2.bib
vignette: >
  %\VignetteIndexEntry{Introduction to splines2}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
output: rmarkdown::html_vignette
---


The package **splines2** is now designed to be a complementary package on
splines.  It provides functions constructing a variety of spline bases that are
not available from the package **splines** shipped with **R**.  Most functions
have a very similar user interface with the function `bs` in package
**splines**. Currently, **splines2** provides function constructing B-splines,
monotone splines (M-splines) and its integral (I-splines), convex splines
(C-splines), and integral of B-splines.  Compared with package **splines**,
**splines2** allows piecewise constant basis for B-splines and M-splines.


In this vignette, we introduce the basic usage of the functions provided by
examples. The details of function syntax and attributes bundled with objects
returned are available in the package manual.


An outline of the remainder of the vignette is as follows: We first introduce
the functions constructing the monotone splines ([M-splines](#mSpline)), its
integral ([I-splines](#iSpline)), and convex splines ([C-splines](#cSpline)).
After then, toy examples for [integral of B-splines](#ibs)
and [B-splines](#bSpline) allowing piecewise constant are given.  Last but not
the least, handy methods of S3 generic function [predict](#predict) for objects
produced by **splines2** are demonstrated for the evaluation of the same spline
basis at new values.


## M-splines using `mSpline`{#mSpline}

M-splines [@ramsay1988monotone] can be considered as a normalized version of
B-splines with unit integral within boundary knots.  An example given by
@ramsay1988monotone was a quadratic M-splines with three internal knots placed
at 0.3, 0.5, and 0.6. The boundary knots by default are the range of the data
`x`, thus 0 and 1 in this example.


```{r mSpline, fig.width=7, fig.height=4, fig.cap = "Quadratic M-splines with three internal knots."}
library(splines2)
knots <- c(0.3, 0.5, 0.6)
x <- seq(0, 1, 0.01)
msOut <- mSpline(x, knots = knots, degree = 2, intercept = TRUE)
library(graphics) # attach graphics just in case for plots
par(mar = c(2.5, 2.5, 0, 0), mgp = c(1.5, 0.5, 0))
matplot(x, msOut, type = "l", ylab = "y")
abline(v = knots, lty = 2, col = "gray") # mark internal knots
```


## I-splines using `iSpline` {#iSpline}

I-splines [@ramsay1988monotone] are simply the integral of M-splines and thus
monotonically non-decreasing with unit maximum value. A monotonically
non-decreasing (non-increasing) function can be fitted by a linear combination
of I-spline bases with non-negative (non-positive) coefficients, plus a constant
function (where the coefficient of the constant function is unconstrained).

The example given by @ramsay1988monotone was the I-splines corresponding to that
quadratic M-splines with three internal knots placed at 0.3, 0.5, and 0.6. Note
that the degree of I-splines is defined from the associated M-splines instead of
their own polynomial degree.


```{r iSpline, fig.width=7, fig.height=4, fig.cap = "I-splines of degree two with three internal knots."}
isOut <- iSpline(x, knots = knots, degree = 2, intercept = TRUE)
par(mar = c(2.5, 2.5, 0, 0), mgp = c(1.5, 0.5, 0))
matplot(x, isOut, type = "l", ylab = "y")
abline(h = 1, v = knots, lty = 2, col = "gray")
```

In additional, the corresponding M-spline basis matrix is shipped in the
attribute named `msMat` in the object returned by function `iSpline`. In other
words, if we need both M-spline bases and their integral splines in model
fitting, extra function call on `mSpline` should be avoided.

```{r msMat}
all.equal(msOut, attr(isOut, "msMat")) # 'msMat' attribute for M-splines
```


## C-splines using `cSpline` {#cSpline}

Convex splines [@meyer2008inference] called C-splines are a scaled version of
I-splines' integral with unit maximum value.  @meyer2008inference applied
C-splines to shape-restricted regression analysis. The monotone property of
I-spines ensures the convexity of C-splines. A convex regression function can be
estimated using linear combinations of the C-spline bases with non-negative
coefficients, plus an unrestricted linear combination of the constant function
and the identity function $g(x)=x$. If the underlying regression function is
both increasing and convex, the coefficient on the identity function is
restricted to be nonnegative as well.


Function `cSpline` provides argument `scale` specifying whether scaling on
C-spline bases is required.  If `scale = TRUE` by default, each C-spline basis
is scaled to have unit height at right boundary knot. The corresponding I-spline
(M-spline) bases shipped in attributes `isMat` (`msMat`) are also scaled to the
same extent and thus serve as the first (second) derivatives of the C-spline
bases. If `scale = FALSE`, the actual integral of I-spline basis will be
returned.  The original I-splines integrated and the corresponding M-splines are
available in attributes named `isMat` and `msMat`, respectively.


```{r cSpline-scaled, fig.width=7, fig.height=4, fig.cap = "C-splines of degree two with three internal knots."}
csOut1 <- cSpline(x, knots = knots, degree = 2, intercept = TRUE)
par(mar = c(2.5, 2.5, 0, 0), mgp = c(1.5, 0.5, 0))
matplot(x, csOut1, type = "l", ylab = "y")
abline(h = 1, v = knots, lty = 2, col = "gray")
```


## Integral B-splines using `ibs` {#ibs}

A close-form recursive formula of B-spline integral given by @de1978practical is
implemented.  One toy example is given as follows:

```{r ibs, fig.width=7, fig.height=4, fig.cap="Piecewise linear B-splines (left) and their integrals (right)."}
ibsOut <- ibs(x, knots = knots, degree = 1, intercept = TRUE)
par(mar = c(2.5, 2.5, 0, 0), mgp = c(1.5, 0.5, 0), mfrow = c(1, 2))
matplot(x, attr(ibsOut, "bsMat"), type = "l", ylab = "y")
abline(v = knots, h = 1, lty = 2, col = "gray")
matplot(x, ibsOut, type = "l", ylab = "y")
abline(v = knots, h = c(0.15, 0.2, 0.25), lty = 2, col = "gray")
```

## B-splines using `bSpline` {#bSpline}

Function `bSpline` provides B-spline bases and allows `degree = 0` for piecewise
constant bases, which is one simple but handy extension to function `bs` in
package **splines**. (For positive `degree`, `bSpline` internally call `bs` to
do the hard work.) Step function or piecewise constant bases are often used in
practice for a reasonable approximation without any assumption on the form of
target function. One simple example of B-splines and M-splines of degree zero is
given as follows:


```{r bSpline, fig.width=7, fig.height=4, fig.cap="B-splines (left) and M-splines (right) of degree zero"}
bsOut0 <- bSpline(x, knots = knots, degree = 0, intercept = TRUE)
msOut0 <- mSpline(x, knots = knots, degree = 0, intercept = TRUE)
par(mar = c(2.5, 2.5, 0, 0), mgp = c(1.5, 0.5, 0), mfrow = c(1, 2))
matplot(x, bsOut0, type = "l", ylab = "y")
abline(v = knots, lty = 2, col = "gray")
matplot(x, msOut0, type = "l", ylab = "y")
abline(v = knots, lty = 2, col = "gray")
```


## Evaluation on New Values by `predict` {#predict}

The methods for **splines2** objects dispatched by generic function `predict`
are useful if we want to evaluate the spline object at possibly new $x$
values. For instance, if we need evaluate the value of I-splines object in
previous example at 0.275, 0.525, and 0.8, respectively, all we need is

```{r predict}
predict(isOut, c(0.275, 0.525, 0.8))
```

Technically speaking, the methods take all information needed, such as `knots`,
`degree`, `intercept`, etc., from attributes of the original **splines2**
objects and call the corresponding function automatically for those new $x$
values. Therefore, the `predict` methods will be not applicable if those
attributes are somehow lost after certain operation.



## Reference

